---  
title: "Course Schedule and Readings"
---  
  While the topics covered are broad, each week will feature different examples from genetics, ecology, molecular, and evolutionary biology highlighting uses of each individual set of techniques.    
  
This course will be a mixture of lecture, live-code demonstrations, and opportunities for in class work. Lecture days will have small exercises for students at the end of class. We will conduct lectures and labs in a computer lab in order for students to be able to follow along and try out new concepts once described and demonstrated in lecture, enabling rapid feedback between students and faculty.
  
W&G = [Wickham & Grolemund](http://r4ds.had.co.nz), DC = Data Carpentry Lesson, U/P for linked pdfs = biol355
    
```{r dates, echo=FALSE, message=FALSE, warning=FALSE}
library(lubridate)
current_date = parse_date_time("2020-01-27", orders="ymd")-days(7)
```
### Week 1. Overview of Course, Data Creation, and Spreadsheets  
```{r next_date, echo=FALSE }
current_date <- current_date+ days(7)
datestring <- paste(month(current_date), day(current_date), year(current_date), sep="/")
```
`r datestring`  
__Lecture:__ [Welcome!](./Lectures/01_intro.pdf),  [Data Collection and Metadata](./Lectures/02_data_creation.html)  
__Readings:__ [W&G Introduction Chapter 1](http://r4ds.had.co.nz/introduction.html), [Browman and Woo 2017](./Readings/Browman_and_Woo_Spreadsheets.pdf)  
__Objective(s):__ Introduce the students to the course; understand what is data, discuss how we preserve information about data, view different examples of datasets from different disciplines. Compare poor versus good practice in creating data. Differentiate between data recording and data entry.  
__Lab:__ [DC Spreadsheet Ecology Lesson](http://www.datacarpentry.org/spreadsheet-ecology-lesson/)  
__Files:__ [Portal Data](https://ndownloader.figshare.com/files/2252083)   
__Etherpad:__ https://etherpad.wikimedia.org/p/355-spreadsheets-2020   

<!--
### Week  2. Introduction to R and RMarkdown  

`r datestring`  
__Readings:__ W&G On [Workflow Basics](https://r4ds.had.co.nz/workflow-basics.html) and [Projects](https://r4ds.had.co.nz/workflow-projects.html)    
__Lectures:__  [Intro to R](./Lectures/04_r_intro.html), [Intro to vectors, lists, matrices, and data frames](./Lectures/05_r_intro_list_matrix_df.html), [Data import and package installation](./Lectures/06_read_data_libraries.html)    
__Objective(s):__ Begin to learn the R computing language. Identify the syntax of an R function (name and arguments); Create an R project in RStudio.  
__Biological Examples:__ Transgenic mouse experimentation. Bee genomics. Subtidal field surveys. Current data sharing archives in biology.
__Etherpad:__  https://etherpad.wikimedia.org/p/biol355-r-intro-2020    

-->
  
### Week 2 Introduction to R and RMarkdown
```{r next_date, echo=FALSE}
```
`r datestring`  
__Readings:__ W&G On [Workflow Basics](https://r4ds.had.co.nz/workflow-basics.html) and [Projects](https://r4ds.had.co.nz/workflow-projects.html), [Tibbles & Data Frames](http://r4ds.had.co.nz/tibbles.html) and [Data Import](http://r4ds.had.co.nz/data-import.html)  
__Lectures:__  DC Ecology [Intro to R](http://www.datacarpentry.org/R-ecology-lesson/01-intro-to-r.html) and [starting with data](http://www.datacarpentry.org/R-ecology-lesson/02-starting-with-data.html)   
__Objective(s):__  Begin to learn the R computing language. Identify the syntax of an R function (name and arguments); Create an R project in RStudio. Read data into R using read.csv(); Use R as a basic calculator; Describe and create variables in R; Interpret the output of the str() function; Install packages in R;  
__Biological Examples:__ Human genome size.   
__Lab:__ [Intro to RMarkdown](https://rmarkdown.rstudio.com/lesson-1.html), [R Markdown and Data Subsetting](./Labs/Lab_Rmarkdown_Subsetting.html)  
__Excercises:__ [Warmup Excercises for Friday](./Labs/lab_2_warmup.html)  
__Cheat Sheets:__ [RMarkdown](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf), [Data Import](https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf)  
__Files:__ [Excel](./Data/my_data.xlsx), [CSV](./Data/my_data.csv)     
__Homework:__ [Data frames!](./Assignments/2_data_frames.html)   
__Etherpad:__  https://etherpad.wikimedia.org/p/biol355-r-intro-2020    

### Week 3. Visualization 
```{r next_date, echo=FALSE}
```
`r datestring`  
__Readings:__ W&G on [Data Visualization](http://r4ds.had.co.nz/data-visualisation.html),  [Unwin 2008](./Readings/Unwin_2008_dataviz.pdf), [Choosing a Good Chart Cheat Sheet](./Readings/choosing-a-good-chart-09.pdf)  
__Lectures:__ [Principles of Data Visualization](./Lectures/07_data_viz_principles.pdf), [Intro to ggplot2](./Lectures/08_ggplot_intro.html)  
__Objective(s):__  Develop understanding of graphical presentation best practices. Create a scatterplot using ggplot(); Learn how to add data to a simple map    
__Biological Examples:__  [Plum Island LTER Plankton Distribution](./Lectures/data/EST-PR-PlanktonChemTax.xls). [HAD CRUT Global temperature anomaly over the past century](./Lectures/data/hadcrut_temp_anomoly_1850_2015.csv)  
__Cheat Sheets:__ [Data visualization](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf), [Mapping with ggmap](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/ggmap/ggmapCheatsheet.pdf)  
__Other References:__[ggplot2 references](http://ggplot2.tidyverse.org/reference/), [Fundamentals of Data Visualization](http://serialmentor.com/dataviz/), [colors for data viz](https://lisacharlotterost.github.io/2016/04/22/Colors-for-DataVis/)  
__Etherpad:__ https://etherpad.wikimedia.org/p/355-dataviz-2020  
__In Class Files:__ [ggplot and hadcrut](./in_class_files/2018/scripts/hadcrut_ggplot.R)  
__Homework:__ [Plotapalooza](./Assignments/3_ggplot_intro.html), submit [here!](https://www.dropbox.com/request/9whCDDUB8TNByJ5KbkoP)

### Week  4. Data Reduction and Summarization
```{r next_date, echo=FALSE}
```

`r datestring`  
__Readings:__ W&G on [Data Transformation](http://r4ds.had.co.nz/transform.html), [Exploratory Data Analysis](http://r4ds.had.co.nz/exploratory-data-analysis.html), and [Pipes](http://r4ds.had.co.nz/pipes.html), DC Ecology Lesson on [Data Aggregation](http://www.datacarpentry.org/R-ecology-lesson/03-dplyr.html)  
__Lecture:__ [Organizing Data to Tell a Story](./Lectures/09_summarizing_data.html)   
__Classwork:__ [Submit here!](https://www.dropbox.com/request/cNqOhHN02iYpldf7KVnQ)  
__Lab:__ [Gapminder and Dplyr](https://swcarpentry.github.io/r-novice-gapminder/13-dplyr/), [Dplyr Faded Example Practice](./Lectures/dplyr_faded_examples.html)<!--, [Data for Exploratory Data Analysis of Mortality and Natality with dplyr and ggplot2](./Labs/freeform_dplyr.zip)-->   
__Cheat Sheets:__ [Data Wrangling Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)  
__Objective(s):__ Describe the meaning and identify applications of the following summary/descriptive statistics: mean, mode, median, standard deviation; Describe the split-apply-combine strategy of data reduction and summarization; Use group_by() and summarise() to calculate summary statistics for groupings within a dataset; Subset data using filter()  
__Biological Examples:__ Human genome size. [Sockeye salmon sizes](./Lectures/data/salmon_rivers.csv).   
__In Class Files:__ [gapminder and dplyr](./in_class_files/2018/scripts/dplyr_lab.R)  
__Homework:__ [Birthweights in California](./Assignments/4_dplyr_assignment.html)  
__Etherpad:__ https://etherpad.wikimedia.org/p/355-dplyr-2020  


### Week  5. Date Cleaning: Strings, Factors, and Dates
```{r next_date, echo=FALSE}
```

`r datestring`  
__Readings:__ W&G on [Strings](http://r4ds.had.co.nz/strings.html), [Dates and Times](http://r4ds.had.co.nz/dates-and-times.html), and [Factors](http://r4ds.had.co.nz/factors.html)  
__Lecture:__  [Strings and Regular Expressions 1](./Lectures/11_regexp.html), [Strings and Regular Expressions 2](./Lectures/12_regexp_2.html)    
__Cheat Sheets:__ [Work with Strings Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf)  
__Objective(s):__ Understand how strings differ from numbers. Learn the basics of string manipulation. Describe the different strategies to clean data full of errors with minimal effort. Process and understand the concept of regular expression matching. Manipulate different date formats and work them into a data map reduce workflow.  
__Files:__ [Portal Mammal Data with String Problems](./Lectures/data/portal_messy_regexp.xlsx)  
__Homework:__ [Regular Expressions](./Assignments/6_regex_practice.html)   
__Etherpad:__ https://etherpad.wikimedia.org/p/355-strings-2020  


### Week  6. Tidy Data and Data Cleaning
```{r next_date, echo=FALSE}
```
`r datestring`  
__Readings:__ W&G Chapters on [Tidy Data](http://r4ds.had.co.nz/tidy-data.html)  
__Lecture:__ [Tidy Data](./Lectures/13_data_tidy.html)  
__Lab:__ [Tidy Coronavirus](./Labs/coronavirus.html)<!-- [Axoltl Data Cleaning](./Labs/axoltl_lab.html)-->  
__Objective(s):__ Understand how to reshape and manipulate data. Describe the difference between the two fundamental forms of data – long versus wide, Use the tidyr package in R to convert between long and wide data; Use unite and separate to create tidy data (where each column is a variable).   
__Biological Examples:__ [Axoltl limb regeneration](./Lectures/data/ALM assay statistics.xlsx). [Mammal taxonomic records](./Lectures/data/mammals.R).  [Weather data](./Data/weather_messy.csv).  [Sale prices for homework](./Data/Sale_Prices_City.csv). [HAD CRUT Global temperature anomaly over the past century, wide format](./Data/hadcrut_temp_anomoly_wide_1850_2015.csv).      
__Etherpad:__ https://etherpad.wikimedia.org/p/355-tidy-2020  
__Homework__: [Weather Data Cleaning](./Assignments/tidyr_over_break_shortened.html), [submit here](https://www.dropbox.com/request/ruza0zXvyvnJgllJXqJ4)    
__MIDTERM__: [make a cheatsheet](./midterm/midterm.html)      

### Week  7 & 8. Data "Mashups" and Geospatial Data

```{r next_date_2, echo=FALSE}
current_date <- current_date + days(7)
datestring_1 <- paste(month(current_date), day(current_date), year(current_date), sep="/")
current_date <- current_date + days(7)  + days(7) 
datestring <- paste(month(current_date), day(current_date), year(current_date), sep="/")
```
`r datestring_1` and `r datestring`  
__Readings:__   [Intro to Geocomputation in R](https://geocompr.robinlovelace.net/intro.html), [Geographic Data in R](https://geocompr.robinlovelace.net/spatial-class.html), W&G Chapter on [Relational Data](http://r4ds.had.co.nz/relational-data.html)    
__Lectures:__  [Intro to GIS](./Lectures/15_gis_intro.html), [Rasters](./Lectures/16_raster_gis_intro.html), [GIS in R](./Lectures/17_gis_in_r.html),  [Joins](./Lectures/15_joins.pptx), [Interactive Maps](./Lectures/18_interactive_maps.Rmd)   
__Labs:__ [Lessons 1,2, 6-8 from Data Carpentry Geospatial Lab](https://datacarpentry.org/r-raster-vector-geospatial/), [coronavirus spatial mapping](./Labs/coronavirus_spatial.html) [joins and maps](./Labs/join_maps_lab.html)  
__Objective(s):__ Know when and where to use different types of joins, Understand how to merge survey data with geospatial information to get a geographic understanding of epidemiological patterns  
__Biological Examples:__ Hemlock wooly adelgid distribution. CDC records of heart disease across counties of the US. Change in coastal sea surface temperature since 1850. Global TB mortality distribution.  
__Optional Reading:__ [Making maps in R](https://geocompr.robinlovelace.net/adv-map.html) and other chapters in Geocomputation in R. [Spatial Data Science](https://keen-swartz-3146c4.netlify.com/), [information about the lab data](https://datacarpentry.org/geospatial-workshop/data/)   
__Files:__  [geospatial data for lab](https://ndownloader.figshare.com/articles/2009586/versions/10), [Hemlock](../Data/hemlock.xlsx), [hemlock_densities](../Data/hemlock_densities.xlsx),  [Arctic Boreal Forest Vegetation](./Lectures/data/boreal.csv), [Heart Disease in America](./Lectures/data/hd_all.xlsx), [data for lab on joins](./Labs/data/joinmap_data.zip), [US County Borders in 2013](./Lectures/data/cb_2013_us_county_500k.zip), [March 18, 2018 SST Anomalies ](./Lectures/data/sst_anom.bin.20180318.tif)   
__Etherpad:__ https://etherpad.wikimedia.org/p/355-gis-2020  
__Homework:__ [Map making!](./assignments/mapping_assignment.html)  
__Before Class:__ Install `rgdal` (you'll need to install gdal first - see below), `sf`, `sp`, `raster`, `leaflet`, `maptools`, `mapdata`, and `rgeos`. 

**To install gdal on a mac, there are two steps**  
1) Install Homebrew from http://brew.sh/ (this is an awesome thing to have anyway)  
2) in Terminal type  
`brew install gdal`  
  
**To install on a Windows PC**  
1) Install OSGEO4W https://trac.osgeo.org/osgeo4w/wiki  
2) Use it to install gdal  

### Week  9 Functions 
```{r next_date, echo=FALSE}
```

`r datestring`  
__Readings:__ W&G Chapters on [Functions](http://r4ds.had.co.nz/functions.html) and [Iteration](http://r4ds.had.co.nz/iteration.html)  
__Lectures:__ [Intro to Functions](./Lectures/19_functions_intro.html), [Iteration with purrr](./Lectures/20_iteration.html)  
__Lab:__ [Functions and Iteration](./Labs/functions_iteration.html)  
__Objective(s):__ Learn the benefits of reusable code, Understand the structure of a function, Discover debugging and making functions fail usefully, Apply conditional logic to build flexible code, Derive principles to make functions that are easy to understand and apply to multiple data sets.  
__Biological Examples:__ NOAA buoy data.  
__Files:__   [NOAA buoy data from Boston Harbor](./Lectures/data/buoydata.zip), [get_buoy.R](./Lectures/get_buoy.R)  
__Homework:__ [Functions!](./Assignments/function_homework.html)  
__Etherpad:__ https://etherpad.wikimedia.org/p/355-functions-2020  

### Week  10. Shiny and Web Interfaces  
```{r next_date, echo=FALSE}
```
`r datestring`
__Readings:__ [Mastering Shiny](https://mastering-shiny.org/) Ch. 1-3, 4 optional, but recommended    
__Objective(s):__ Learn how to communicate data to others using dynamic web based applications.  
__Lectures:__ [Building Shiny Apps](https://docs.google.com/presentation/d/1dXhqqsD7dPOOdcC5Y7RW--dEU7UfU52qlb0YD3kKeLw/edit#slide=id.gd565d6fc7_0_182)  
__Lab:__ [Lab for Building Shiny Apps](https://deanattali.com/blog/building-shiny-apps-tutorial/)  
__Data:__[download](http://pub.data.gov.bc.ca/datasets/176284/BC_Liquor_Store_Product_Price_List.csv)  
__Additional Resource:__ https://shiny.rstudio.com/tutorial/  
__Biological Examples:__ Covid-19 Interactive data exploration.  
__Etherpad:__ [Shiny homework with coronavirus](./Labs/shiny_lab.html)  
__Etherpad:__ https://etherpad.wikimedia.org/p/355-shiny-2020  


### Week  11. T-Tests and P-Values 
```{r next_date, echo=FALSE}
```
`r datestring`  
__Readings:__ [Cortina and Dunlop 1997](./Readings/cortina and Dunlop 1997 significance testing p value.pdf)  
__Lectures:__ [Introduction to Frequentist Hypothesis Tests](./Lectures/21_testing_hypotheses_p.pdf), [T-Tests](./Lectures/22_t.pdf)  
__Objective(s):__ Describe the basics of probability and p-values, Compare groups of data using T-tests and its extensions.  
__Biological Examples:__ The effects of testosterone on bird behavior.  
__Lab:__ [T-Tests](./Labs/lab_9_t.html)  
__Files:__ [Blackbird Testosterone Implants](./Labs/data/12e2BlackbirdTestosterone.csv), [Survivorship of Salmon and Trout](./Labs/data/12e4BrookTrout.csv)  
__Etherpad:__ https://etherpad.wikimedia.org/p/355-ttest-2020  
__Homework:__ [t-tests](./Assignments/9_t.html), [Lizard data](./Lectures/t_test/12e3HornedLizards.csv)  
  
### Week 12. Analysis of Variance and the Linear Model
```{r next_date, echo=FALSE}
```
 `r datestring`  
__Lectures:__ [Analysis of Variance](./Lectures/23_anova.html), [Linear Models](./Lectures/24_lm.html)    
__Lab:__ [Linear Models and ANOVA](./Labs/lm_anova_lab.html), [data files for lab](./Labs/data/lm_lab_data.zip)  
__Objective(s):__ Analyze the differences between multiple means.  Fit a linear regression using lm() in R through a bivariate scatterplot. Evaluate assumptions of a linear model with normal error   
__Biological Examples:__  Jet-lag recovery and circadian rhythms. Seal life history variation  [files](./Lectures/data/anova.zip)  
__Homework:__ [Linear temperature change?](./Assignments/10_lm_anova.html)  
__Etherpad:__ https://etherpad.wikimedia.org/p/355-anova-2020   

### Week  13  General and Generalized Linear Models, Multiple Predictors  
```{r next_date_2, echo=FALSE}
```
`r datestring_1`  
__Readings:__  W&G on [Model Basics](http://r4ds.had.co.nz/model-basics.html)   
__Lectures:__ [The General Linear Model (ANCOVA, multiple regression)](./Lectures/26_ancova_mulltreg.html), [Generalized Linear Models](./Lectures/25_glm.html), [AIC and Multimodel Comparisons](./Lectures/27_mmi.html)  
__Lab:__ [Linear Models](./Labs/lab10_lm.html)  
__Objective(s):__ Describe when to use nonlinear models/curves, Comparing and Contrasting models, Visualization of model outcomes  
__Biological Examples:__ Seal life history variation. Mouse anti-fungal drug development. Neanderthal brain size. Fire severity in California.   
__Files:__[Files for lab](./Labs/data/lm_data.zip)  
__Etherpad:__ https://etherpad.wikimedia.org/p/355-mlr-2020

<!--
### Week  14 Multiple Predictors Part 2  
`r datestring`  
__Readings:__  W&G on [Model Building](http://r4ds.had.co.nz/model-building.html)  
__Lectures:__  [AIC and Multimodel Comparisons](./Lectures/27_mmi.html), [purrr](./Lectures/purrr.html).    
__Lab:__ [AIC and Multimodel Inference in R](./Labs/lab11_mmi.html)  
__Objective(s):__  Comparing and Contrasting models, Visualization of model outcomes when many models are fit  
__Biological Examples:__ Fire severity in California. Brain size and natal investment in primates. Linguistic diversity and cultural characteristics.   
__Files:__ [Files for lab](./Labs/data/mmi_data.zip)  
__Etherpad:__ https://etherpad.wikimedia.org/p/355-aic-2020

-->

### Week  14. Open Lab and Final Presentations
```{r next_date, echo=FALSE}
```
`r datestring`  
Prep your final talks! Data science party!