---
title: "First Intro to R"
author: "Intro to Data Science"
output: html_document
---
<!-- for next year, add NULL, INF, exercise at end? -->

Hello!  So, today we're going to begin to code in R. We're going to cover the basics of using R with a focus on data frame objects.

But let's begin. To talk about what's going on offline, use today's etherpad at https://etherpad.wikimedia.org/p/biol355-r-intro

### File structure
For this and all in class code exercises, I want you to save what you do in a file. To do that, open a new file. Awesome. Now, before we do anything, I want to talk about commenting. In R, we can write comments. Comments are never interpreted. So, for example, put the following in your file, and then copy and paste it (or hit cmd-enter or ctrl-enter) into your console. Notice nothing happens.  

```{r a_comment}
#Nothing is going to happen
```

Comments are terribly useful, because you can use them to add information to a file. Often, before I start a script, I'll lay it out in a few chunks. First, I'll supply header information. Second, I'll write out my workflow in plain english comments. For example:

```{r more_comments}
###########################
#' my_first_file.R
#' 
#' This script is my first R file
#' and it doesn't do much, but it's mine.
#'
#' @author Jarrett Byrnes
#'
#' @changelog
#' 2018-02-01 Added a changelog
#' 2018-01-31 finished this file
###########################

## Do some things here ####

## Do some other things here ####

```

A few more advanced things to notice here. First, in the header, I have `#'` a lot. This is a convention in R such that whenever you add a linebreak, you'll start off with more comments on the next line. There also some `@` stuff. There is actually a standard vocabulary here, but for now, know that when you do that in `#'` comments, it gets highlighted, and is a nice way to help delimit different fields in your headers.  
  
Second, in the blocks of comments after the header, I used a `####`. This is a really nice feature for Rstudio where then whatever was before that is recognized as a section header by RStudio. It will make your file easier to work with later.

**Exercise:** Setup a script file for today!  

OK, that's comments - now, on to the code!

###I have this cursor sitting here. What can I do with it?

Peering into your console, often the first source of fear and confusion is, what the heck do I do with this giant blank space.  The easiest way to start thinking about R is as th world's most advanced calculator.  Seriously, try it out!

```{r}
3+4
```

Whoah!  You can add!  Wonderful.  Let's try a few other operations.

```{r}
3-4

4*5

9/3

2^7
```

Yes, basic arithmatic is right there at your fingertips.

**Exercise**: Try some other basic arithmetic operators. Anything not work the way you think it should?

###Arithmetic is great. But I want more

While arithmetic is cool (and necessary) we often want R to do more for us. Perhaps calculating more compelling mathematical functions. One that we commonly use is logarithms. Let's say you want the natural log of 10.

```{r}
log(10)
```

Great! Notice how that worked. We had log. Then two parentheses. Inside of which we had 10.  log is a **function**. Functions in R are hugely powerful. They are the core of many things we do. Functions consist of a function name, those parentheses, and then inside of the parentheses one or more *arguments* separated by commas. Often these arguments have names. For example, what if we wanted to get the log of 10 in base 10.

```{r}
log(10, base = 10)
```

Now we've supplied two arguments.


###Help! I don't know what arguments to give to a function!

We're going to run into many different functions as we go forward. To get help on how to use them, and what arguments to supply, there are two ways to get help. Let's look at the help file for log:

```{r eval=FALSE}
?log

help("log")
```

Both of those do the same thing. Note that there is an order to arguments in the help file. If you put arguments in order in a function, you don't need to worry about naming them. This is bad practice (future you will be unhappy), as you may forget what they mean. Whevenever possible, use named arguments.


###But I don't know what the name of a function I need is!

Sometimes, we don't know the name of a function we're looking for. For example, in R, the function for arcsin is `asin`. But you have no way of knowing that. For that, we use the ?? with quotes.

```{r eval=FALSE}
??"arcsine"
```

Notice this brings up a list of helpfiles that have the word arcsine in them. Bueno! You can now track down the right function.

**Exercise**: Think of a simple mathematical function. Find its helpfile and then implement it in R.

###Before we go any further, a comment on comments

It can be VERY easy to get lost in a sea of R code, not knowing what is going on. Fortunately, R provides something called *comments*.  In a comment, R stops evaluating code, and let's you write whatever love notes to yourself that you want to write.  In R the comment character is `#`.  For example.

```{r}
#this is a comments

### This is also a comment

3+4 #hey, I commented after 3+4

########################################
##### Oh, a comment box
##### that I can use to delinieate
##### blocks of code
########################################
```

**ALWAYS COMMENT THE HECK OUT OF YOUR CODE TO HELP OUT FUTURE YOU!**  I'll include some comments in the code today to show you examples.

###Variables and You

One of the great things about R is that you can save things as variables and use them later on. Some of them are there already.  For example:

```{r}
#This is Pi
pi
```

WHOAH! PI IN R! 

What if you wanted to make your own variable. Say, foo. And you wanted foo to always equal the square root of 2.

```{r}
#let's create a variable foo
foo <- sqrt(2)


#what's inside of foo
foo
```

Note that the assignment operator is <- and not an = sign. Now, you *can* use =, but in R it's generally bad practice, as = will crop up in other places, and you'll want to avoid it.

Now you have a variable that you can use form now on that is the square root of two! For example:

```{r}
foo + 5

log(foo, base=2)

foo^2
```

###More than a number

OK, numbers are great, but there are other types of objects we'll be dealing with in R. Primarily, we're going to work with data frames, but let's build up to a data frame, as it's big and hairy. 

First, are there other object types that have a single element to them beyond numbers?  Well, yes!  There are strings - words in quotes

```{r}
"hello"
```

Also, Booleans, which denote true and false
```{r}
TRUE

FALSE

0==4

4==4

3 <= 4

3 <= 4
```

Note the different ways we made comparisons.  These will become handy as you move on. Booleans are really 1s and 0s, such that you can even do math with them

```{r}
TRUE + TRUE + FALSE
```

OK, what about something for a missing value. For that, we have NA.  This is quite important if, say, you have a missing value in your data set.

```{r}
NA

NA + 1
```

No one likes an NA, and often we have to find ways around them!

