---
output: html_document
---

Learning objectives:

1. Define statistics.

2. Explain the difference between a sample and a population.

3. Compare accuracy and precision

4. Describe a random sample and provide an example of a random sampling design.

5. Plot a frequency distribution given a dataset.


```{r}
library(readr)
library(ggplot2)

salmon <- read_csv('../Data/02e1cSockeyeFemaleBodyMass.csv')

number_ticks <- function(n) {function(limits) pretty(limits, n)}

p <- ggplot(data = salmon, aes(x = mass)) + 
  geom_dotplot(binwidth = 0.05) + 
  coord_fixed(ratio = 0.05) + 
  ylim(0, 20) + 
  scale_x_continuous(breaks=number_ticks(20)) + 
  theme(axis.text = element_text(size = 24), 
        axis.title = element_text(size = 32))
p

mass_range <- data_frame(range = paste0('range = ', min(salmon$mass), ' - ', max(salmon$mass)), 
                         mean  = paste0('mean = ', round(mean(salmon$mass), 2)))

p + geom_vline(xintercept = mean(salmon$mass), colour = 'red', size = 2) + 
    geom_text(data = mass_range, aes(x = 2.2, y = 18, label = mean), size = 14, hjust = 0)

p + geom_vline(xintercept = mean(salmon$mass), colour = 'red', size = 2) + 
    geom_text(data = mass_range, aes(x = 2.2, y = 18, label = mean), size = 14, hjust = 0) +
    geom_text(data = mass_range, aes(x = 2.2, y = 15, label = range), size = 14, hjust = 0)

salmon$intervals <- cut_width(salmon$mass, width = 0.05)

```

```{r, eval = FALSE}
salmon_intervals <- 
  dplyr::group_by(salmon, intervals) %>% 
    dplyr::summarise(frequency = length(intervals)) %>% 
    as.data.frame()

write.csv(salmon_intervals, '09_salmon_intervals.csv')
```



```{r}
number_ticks <- function(n) {function(limits) pretty(limits, n)}

p <- ggplot(data = salmon, aes(x = mass)) + 
  geom_dotplot(binwidth = 0.05) + 
  coord_fixed(ratio = 0.05) + 
  ylim(0, 20) + 
  theme_bw() +
  scale_x_continuous(breaks=number_ticks(20)) + 
  theme(axis.text = element_text(size = 24), 
        axis.title = element_text(size = 32))
p


d <- ggplot(data = salmon, aes(x = mass)) + 
  geom_density(fill = 'blue') +
  theme_bw() +
  scale_x_continuous(breaks=number_ticks(20)) + 
  theme(axis.text = element_text(size = 24), 
        axis.title = element_text(size = 32)) + 
  ylab('Probability density')
d

ggdraw(switch_axis_position(d, axis = 'y'))

```

```{r}
# Code used at the very end of class
library(ggplot2)

salmon <- read.csv('../Data/02e1cSockeyeFemaleBodyMass.csv')

salmon

ggplot(data = salmon, aes(x = mass)) +
  geom_histogram(binwidth = 0.05)

ggplot(data = salmon, aes(x = mass)) +
  geom_density(fill = 'blue', alpha = 0.4) +
  geom_vline(xintercept = mean(salmon$mass), colour = 'red', size = 1)
```






Frequency distributions and probability distributions
Frequency = the number of times a specific measurement comes up



Exercise:

What kind of sampling strategy would use to sample the following:




Cat falling from buildings example


1) Describe the meaning and identify applications of the following summary/descriptive statistics: mean, mode, median, standard deviation

