---
title: "Interactive Maps!"
author: "Intro to Data Science for Biology"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### From static maps to dynamic maps

So,  we saw how we could plot shapefiles in ggplot2. This is great! But, oddly, sometimes static. What if you want to no only put shapefiles on a map, but add other information (say, sites), and then make the whole thing interactive - like a google map!  You can!

### Interactive maps with leaflet

Leaflet is a javascript library for making interactive maps. You've probablys een them around before. But there is a nice [r package](https://rstudio.github.io/leaflet/) for them as well that allows you to do a lot! And it works well with the tidyverse

```{r leaflet, message=FALSE}
library(tidyverse)
library(leaflet)

leaflet() %>%
  addTiles()

```

Whoah! Interactive worldmap! With tiles from the OpenStreetMap project! Zoom around and try it. And definitely bookmark https://rstudio.github.io/leaflet/ for a ton of options!


### What about our shapefiles?

Let's revisit the heart disease mortality data, but this time with a shapefile.  Here I've grabbed the 2016 US census county borders from https://www.census.gov/geo/maps-data/data/cbf/cbf_counties.html - let's look at them!

```{r shapefile, message=FALSE}
library(rgdal)

counties <- readOGR("./data/cb_2013_us_county_500k")
```

What's great is that we can make county borders into a leaflet map in a snap!

```{r leaflet_shapefile, cache=TRUE}
leaflet(counties) %>%
  addPolygons(color="black", fillColor = "white", weight=0.1)
```

We can even add some more detail so we know what we're mousing over
```{r mouseover, cache=TRUE}
leaflet(counties) %>%
  addPolygons(color="black", fillColor = "white", weight=0.1,
              highlightOptions = highlightOptions(color = "blue", weight = 2,
              bringToFront = TRUE))
```

We can even put this back into the context of a world map

```{r mouseover2}
leaflet(counties) %>%
  addTiles() %>%
  addPolygons(color="black", fillColor = "white", weight=0.1,
              highlightOptions = highlightOptions(color = "blue", weight = 2,
              bringToFront = TRUE))
```


### Adding heart disease fill data to shapefiles
Let's load in our heart disease mortality data, and see how it matches with the data in our shapefile.

```{r load_heart, message=FALSE}
library(readxl)

heart <- read_excel("./data/hd_all.xlsx",
                    na = "Insufficient Data")
head(heart)

head(counties@data)
```

So, no column names in common. NAME and County **might** match. Let's see!

```{r anti}
bad_match <- anti_join(heart, counties@data,
                       by = c("County" = "NAME"))

nrow(bad_match)

bad_match
```

Not bad! It's just cities! And a few odd counties and the USVI. Works for a join! What's different here is that we're going to join our data INTO the SpatialPolygonsDataFrame.

```{r join}
counties@data <- left_join(counties@data, heart,
                           by = c("NAME" = "County"))
```

So, now we have to plot this. Leaflet has some funky plotting syntax for colors. We need to specify a palette function and use a ~ to say it's being evaluated. I'm also taking away border stroking here.

```{r fill, cache=TRUE, cache=TRUE}
leaflet(counties) %>%
  addTiles %>%
  addPolygons(stroke=FALSE,
             fillColor = ~colorNumeric("magma", Death_Rate)(Death_Rate)) 

```

I find it's a little easier to define my palette before plotting

```{r fill2, cache=TRUE, cache=TRUE}
pal <- colorNumeric("magma", 
                    domain = counties@data$Death_Rate)

leaflet(counties) %>%
  addTiles %>%
  addPolygons(stroke=FALSE, color = ~pal(Death_Rate)) 

```


What does this mean? Well, we want a legend! Fortunately, there's an `addLegend` function!

```{r legend, cache=TRUE}
leaflet(counties) %>%
  addTiles %>%
  addPolygons(stroke=FALSE, color = ~pal(Death_Rate))  %>%
  addLegend(pal = pal, 
            values = ~Death_Rate)

```

This might not be specific enough, so we can add an interactive popup function!
```{r popup, cache=TRUE}
leaflet(counties) %>%
  addTiles %>%
  addPolygons(stroke=FALSE, color = ~pal(Death_Rate),
              popup = ~str_c("Death Rate:", Death_Rate, sep=" "))  %>%
  addLegend(pal = pal, 
            values = ~Death_Rate)

```

### Popup Points!
As a side note, leaflet is also great for points. For example, let's make 200 random locations and put them on a map.

```{r points}
loc_df <- data.frame(x=runif(200, -180, 180), y = runif(200, -100, 100))

leaflet(loc_df) %>%
  addTiles() %>%
  addCircleMarkers(lng = ~x, lat = ~y)
```

There are of course a million options to customize these circles that are similar to polygons. And other similar functions. Try `addMarkers` for fun. Pretty great!

### Exercise

Try and make your own heart disease maps with different palettes and color scales. For reference, here' are the color functions, but see their help files for more details:

- colorNumeric is a simple linear mapping from continuous numeric data to an interpolated palette.

- colorBin also maps continuous numeric data, but performs binning based on value (see the cut function).

- colorQuantile similarly bins numeric data, but via the quantile function.

Each takes a `palette` argument which can be any one of the following:
The palette argument can be any of the following:

1. A character vector of RGB or named colors. Examples: palette(), c("#000000", "#0000FF", "#FFFFFF"), topo.colors(10)

2. The name of an RColorBrewer palette, e.g. "BuPu" or "Greens".

3. The full name of a viridis palette: "viridis", "magma", "inferno", or "plasma".

4. A function that receives a single value between 0 and 1 and returns a color. Examples: colorRamp(c("#000000", "#FFFFFF"), interpolate="spline").



#### Rasters

This is great, but what about adding rasters? No problem! Leaflet has you covered with `addRasterImage` - it even reprojects your raster if its in the wrong projection before adding it!

```{r rast, cache=TRUE, message=FALSE}
library(raster)
rast <- raster("./data/HARV_dsmCrop.tif")

rpal <- colorNumeric("viridis", 
                    values(rast),
                    na.color = "transparent")

leaflet() %>% 
  addTiles() %>%
  addRasterImage(rast,  colors = rpal, opacity = 0.8) %>%
   addLegend(pal = rpal, values = values(rast))
```

### Exercise
Let's take a look at the distribution of global SST anomalies. I've grabbed last week's data from http://www.cpc.ncep.noaa.gov/products/GIS/GIS_DATA/sst_oiv2/sst-anom.php (download it from this site [here](./data/sst_anom.bin.20180318.tif) and I'd like you to load it, and the [marine ecoregions of the world](./data/MEOW-TNC.zip) in order to make a map where we can explore how that temperature anomoly varies by ecoregion. This means you will be layering polygons from a shapefile on top of a raster using leaflet. Feel free to play with color palettes and any other property you'd care to examine!

Note, you'll need to supply the polygons name as the "data" to `leaflet` and supply the raster separately. `fillOpacity = 0` will also make them transparent

Also, include the following line to take out land, etc.
```{r, eval=FALSE}
values(sst)[values(sst)>100] <- NA
```

This sets all of those big values to NA

It should look something like...

```{r test, echo=FALSE, message=FALSE, cache=TRUE}
library(raster)

sst <- raster("./data/sst_anom.bin.20180318.tif")
values(sst)[values(sst)>100] <- NA

meow <- readOGR("./data/MEOW-TNC")

sstpal <- colorNumeric("RdBu", 
                    values(sst),
                    na.color = "transparent")

leaflet(meow) %>%
  addRasterImage(sst, colors = sstpal) %>%
  addPolygons(weight = 1, color="black", fillOpacity = 0)%>%
   addLegend(pal = sstpal, values = values(sst))

```