---
title:
css: style.css
output:
  revealjs::revealjs_presentation:
    reveal_options:
      slideNumber: true
      previewLinks: true
    theme: white
    center: false
    transition: fade
    self_contained: false
    lib_dir: libs
---

## Faded Examples and Dplyr Hurricanes
```{r prep, include=FALSE, warning=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(fig.height=5, fig.width=7, comment=NA, 
               warning=FALSE, message=FALSE, 
               dev="jpeg")

library(rrricanesdata)
library(lubridate)
library(magrittr)
library(dplyr)
library(ggplot2)
states_map <- map_data("world")
theme_set(theme_bw(base_size = 14))
  
adv_tmp <- adv %>%
  mutate(
  Year = year(Date),
  Month = month(Date),
  Day = day(Date),
  Decade = round(Year/10)*10
)
```



```{r show_rrricanes, fig.width=11, echo=FALSE, warning=FALSE, fig.height=3}
ggplot(adv_tmp, aes(x=Lon, y=Lat)) + 
  facet_wrap(~Decade) + 
  geom_line(mapping=aes(alpha=Wind, group=Name, color=Name), lwd=0.1) +
  guides(color="none") +
  geom_path(data = states_map, mapping=aes(x=long, y=lat, group = group)) +
  coord_cartesian(xlim=c(-150,0), ylim=c(8,50))

```

## Libraries for today
```{r libraries, message=FALSE, warning=FALSE}
library(rrricanesdata)
library(lubridate)
library(magrittr)
library(dplyr)
library(ggplot2)
```

Install any you don't have

## Faded Examples...

```{r eval=FALSE}
obj <- old_obj %>%
  ____(newcol = oldcol + 1)
```

## The rrricanesdata
```{r echo=FALSE}
adv
```

## Max Gusts per Storm
```{r max_per_storm}
adv_strength <- adv %>%
  group_by(Name) %>%
  filter(Gust == max(Gust)) %>%
  ungroup()
```

<div id="fragment">The `ungroup()` is only needed if there is no `summarize()`</div>

## Max Forward Speed per Storm
```{r, eval=FALSE}
adv_fwd <- adv %>%
  ____(Name) %>%
  filter(FwdSpeed == ___(FwdSpeed)) %>%
  ungroup() 
```  

```{r, include=FALSE}
adv_fwd <- adv %>%
  group_by(Name) %>%
  filter(FwdSpeed == max(FwdSpeed)) %>%
  ungroup() 
```  

## Minimum Pressure per Storm
```{r eval=FALSE}
adv_min_pressure <- adv %>%
  ___(____) %>%
  ____(Pressure == ____(Pressure)) %>%
  ungroup()
```

```{r include=FALSE}
adv_mean_pressure <- adv %>%
  group_by(Name) %>%
  filter(Pressure == min(Pressure)) %>%
  ungroup()
```

## Maximum Latitude per Storm Type
```{r eval=FALSE}
adv_max_lat <- adv %>%
  ____(____) %>%
  ____(Lat == ____(____)) %>%
  ____()
```

```{r include=FALSE}
adv_max_lat <- adv %>%
  group_by(Status) %>%
  filter(Lat == max(Lat)) %>%
  ungroup()
```

##
<br><br><br>
<h1> Grouping and summarizing</h1>

## Average Gust per Status Only
```{r}
adv_gust_status <- adv %>%
  group_by(Status) %>%
  summarize(Mean_Gust = mean(Gust, na.rm=TRUE))
```

<div class="fragment">How many rows?</div>

## Average Forward Speed by Storm
```{r eval=FALSE}
avg_forward_speed <- adv %>%
  ____(Name) %>%
  summarize(Mean_FwdSpeed = mean(FwdSpeed, na.rm=____))
```

```{r include=FALSE}
avg_forward_speed <- adv %>%
  group_by(Name) %>%
  summarize(Mean_FwdSpeed = mean(FwdSpeed, na.rm=TRUE))
```

## Averge Pressure by Status
```{r eval=FALSE}
avg_pressure_status <- adv %>%
  ____(____) %>%
  ____(Mean_Pressure = ____(Pressure, ____=____))
```

```{r include=FALSE}
avg_pressure_status <- adv %>%
  group_by(Status) %>%
  summarise(Mean_Pressure = mean(Pressure, na.rm=TRUE))
```


##

<br><br><br>
<h2>A little mutate</h2>

## Add "Quarter"
```{r, message=FALSE, warning=FALSE, echo=TRUE}
#a library for dates
#(more on thursday!)
library(lubridate)

adv <- adv %>%
  mutate(Quarter = quarter(Date))
```

## What is Quarter?
```{r echo=FALSE, fig.height=6, fig.width=6}
qplot(month(Date), Quarter, data=adv)
```

## Testing before overwriting

```{r, echo=TRUE}
adv_test <- adv %>%
  mutate(Quarter = quarter(Date))
```

## Testing before overwriting
```{r, echo=TRUE}
adv_test$Quarter[1:100]
```

## Testing before overwriting
```{r, echo=TRUE}
qplot(month(Date), Quarter, data=adv_test)
```

## Testing before overwriting
```{r, echo=TRUE}
unique(adv_test$Quarter)
```

## Now I feel comfortable
```{r eval=FALSE, echo=TRUE}
adv <- adv %>%
  mutate(Quarter = quarter(Date))
```

## Add month, day, and year

```{r, eval=FALSE}
adv <- adv %>%
  ____(Month = month(____),
         Day = day(____),
         Year = year(____)
         )
```

```{r, include=FALSE}
adv <- adv %>%
  mutate(Month = month(Date),
         Day = day(Date),
         Year = year(Date)
         )
```

## Challenge: add decades
```{r, eval=FALSE}
adv <- adv %>%
  ____(Decade = ____(____/____)*____
         )
```

Hint: you might need a nice round function...

```{r, include=FALSE}
adv <- adv %>%
  mutate(Decade = round(Year/10)*10
         )
```

## Plot Average of Max Gusts per Year by Status
```{r eval=FALSE}
adv_gust_status <- adv %>%
  ____(Status, Year) %>%
  ____(Max_Gust = ____(____, ____ = ____))

ggplot(adv_gust_status, mapping=aes(x = ____,
                                    y = ____,
                                    color = ____)) +
  geom_line()
```

```{r maxgust, include=FALSE, eval=FALSE}
adv_gust_status <- adv %>%
  
  group_by(Status, Year) %>%
  summarize(Max_Gust = max(Gust, na.rm=T))

ggplot(adv_gust_status, mapping=aes(x = Year,
                                    y = Max_Gust,
                                    color = Status)) +
  geom_line()
```

## Plot Average of Max Gusts per Year by Status
```{r maxgust, echo=FALSE, warning=FALSE}
```


## 
<Br><br><br>
<h1>Some more complex workflows</h1>

## Plot of Average Monthly Max Gusts per Decade
```{r decadalgust, eval=FALSE}
adv_decadal_monthly_max_gust <- adv %>%
  
  group_by(Decade, Year, Month) %>%
  summarize(Max_Gust = max(Gust, na.rm=TRUE)) %>%
  
  group_by(Decade, Month) %>%
  summarize(Mean_Max_Gust = mean(Max_Gust))


#plot it!
ggplot(adv_decadal_monthly_max_gust, mapping = aes(x = Month,
                                           y = Mean_Max_Gust,
                                           color = factor(Decade))) +
  geom_line()
```

## Plot of Average Monthly Max Gusts per Decade
```{r decadalgust, echo=FALSE}
```

## Plot Average Maximum Storm Windspeeds by Year
```{r eval=FALSE}
adv_storm_annual <- adv %>%

  ____(Name, Year) %>%
  summarize(Max_Wind = max(Wind, na.rm=TRUE)) %>%
  
  ____(Year) %>%
  summarize(Max_Wind_Mean = ____(Max_Wind))


#plot it!
ggplot(adv_storm_annual, aes(x=Year, 
                             y = ____)) +
  geom_line()
```

```{r stormplot, include=FALSE}
adv_storm_annual <- adv %>%

  group_by(Name, Year) %>%
  summarize(Max_Wind = max(Wind, na.rm=TRUE)) %>%
  
  group_by(Year) %>%
  summarize(Max_Wind_Mean = mean(Max_Wind))

ggplot(adv_storm_annual, aes(x=Year, 
                             y = Max_Wind_Mean)) +
  geom_line()
```

## Plot Average Maximum Storm Windspeeds by Year
```{r stormplot, echo=FALSE, warning=FALSE}
```


## Plot the Rank Order Distribution of Gusts by Decade
```{r eval=FALSE}
adv_orders <- adv %>%
  ____(____, Name) %>%
  ____(Max_Gust = max(Gust, ____=____)) %>%

  ____(Decade) %>%
  ____(Rank = rank(desc(Max_Gust))) %>%
  ungroup()


#plot it!
ggplot(adv_orders, mapping=aes(x=____, 
                               y = ____,
                               color=factor(____))) +
  geom_line()
```

```{r rankstorms, include=FALSE}
adv_orders <- adv %>%
  group_by(Decade, Name) %>%
  summarize(Max_Gust = max(Gust, na.rm = TRUE)) %>%

  group_by(Decade) %>%
  mutate(Rank = rank(desc(Max_Gust))) %>%
  ungroup()

ggplot(adv_orders, mapping=aes(x= Rank, 
                               y = Max_Gust,
                               color=factor(Decade))) +
  geom_line()
```

## Plot the Rank Order Distribution of Gusts by Decade
```{r rankstorms, echo=FALSE}
```