---
title:
output:
  revealjs::revealjs_presentation:
    reveal_options:
      slideNumber: true
      previewLinks: true
    theme: white
    center: false
    transition: fade
    self_contained: false
    lib_dir: libs
    css: style.css
---




## 
<Br><Br><br><br>
<h2><center>Make your code purrr</center></h2>


```{r prep, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(stringr)
library(ggplot2)
library(modelr)
theme_set(theme_bw(base_size=16))

opts_chunk$set(fig.height=5, fig.width=7, comment=NA, 
               warning=FALSE, message=FALSE, 
               dev="jpeg")

library(readxl)
library(repurrrsive)
library(listviewer)
```

## load some packages
```{r read_data, eval=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(repurrrsive)
library(listviewer)
```

1. `purrr` is the star today, but it's part of the tidyverse package. If you want to load it alone, `library(purrr)` does the trick.  

2. `repurrrsive` contains example datasets.  

3. `listviewer` is similar to `visdat`. It makes an interactive widget to explore lists, which can be more tricky to grok than the average dataframe.




##Explore lists interactively with listviewer

```{r listviewer}
listviewer::jsonedit(got_chars, mode = "view")
```  
Note: how pretty this is depends on the underlying structure of the list. 

##So what's the deal with lists?  

Most of the functions we've worked with so far are vectorized. They work extremely well on vectors (remember - each column of a dataframe is a vector). 

```{r sqrt}
sqrt(c(9, 16, 25))
```  
However, they break when you try to run them on a list.
```{r sqrt list,  error = TRUE}
a <- list(9, 16, 25)
sqrt(a)
```
## Map it out!  

If you want to apply a function across a list, use `map(list, function)`.

```{r sqrt map}
map(a, sqrt)
```

Note: this works on atomic vectors as well, but the output will be a list. 

##Lists can get biiiiig
`got_char` is a list included in the `repurrrsive` package. It has data on all PoV characters from A Song of Ice and Fire.  

```{r list str}
# str is, as always, a good place to start.
#give str additional arguments in order to manage large lists
str(got_chars, list.len = 3, max.level = 2)
```

##Dust off your index-fu

Find some data about a single character
```{r one}
got_chars[[1]][c("name", "culture", "gender", "born")]
```
Exercise: Who is the last entry in this dataset, and where were they born?  

##Exercise: Who is the last entry in this dataset, and where were they born?
```{r last}
got_chars[[length(got_chars)]][c("name","born")]
```



##What if we want this info for all characters at once? 

```{r []}
x <- map(got_chars, `[`, c("name", "culture", "gender", "born"))
str(x[1]) 
#In english: take the got_chars list. 
#apply the [] function, aka extract() to the entire list. 
#Only spit out name, culture, gender, and birth info.
```
Note: If [ grinds your gears, you can use magrittr's `extract()`.

## Exercise
Pull out the names and titles of each character

## Exercise
Pull out the names and titles of each character
```{r}
y <- map(got_chars, `[`, c("name", "titles"))
str(y[1])
``` 

## Exercise
Pull out the names and titles of each character
```{r}
y <- map(got_chars, `[`, c("name", "titles"))
str(y[1])
y[[1]]$titles
```

## The toolbox
1. map
2. map2
3. pmap
4. map_chr and friends
5. map_df
6. safely()

##map
Lets take a look at map with some numbers.
```{r more map}
a <- rnorm(n = 100, mean = 0, sd =  1)
b <- rnorm(n = 100, mean = 5, sd =  1)
c <- rnorm(n = 100, mean = 10, sd =  1)
d <- list(a, b, c)

map(d, mean)
```

## map2 

If you have two input lists, use map2. 
```{r map2}
a <- list(1, 2, 3)
b<- list(10, 11,12)
#map also works with formulas
map2(a, b, ~ .x + .y)
#take two lists (a and b) 
#add the corresponding elements together
```

##pmap
```{r pmap}
c <- list(100, 101,102)
d <- list(a, b, c)

pmap(d, sum)
```
##Simple Outputs
`map()` will ALWAYS return a list. If you want a simple vector as output, use the appropriate `map` varient from the following list:
1. `map_lgl()` returns a logical (`TRUE`/`FALSE`) vector
2. `map_int()`returns an integer vector
3. `map_dbl()` returns a double vector
4. `map_chr()` returns a character vetor  *

NOTE: `purrr` is *extremely* picky about object type, so be aware of `int` vs `dbl`.

##Simple Outputs
```{r map_chr1}

x <- map(got_chars, `[`, "name")
x[1:2]
```
Versus
```{r map_chr2}
y <- map_chr(got_chars,  ~.x[["name"]])
head(y)
```

##Simple Outputs
```{r map_dbl1}
a <- rnorm(n = 100, mean = 0, sd =  1)
b <- rnorm(n = 100, mean = 5, sd =  1)
d <- list(a, b)
map(d, mean)
```
versus
```{r map_dbl2}
a <- rnorm(n = 100, mean = 0, sd =  1)
b <- rnorm(n = 100, mean = 5, sd =  1)
d <- list(a, b)

map_dbl(d, mean)
```
##map_df

```{r map_df}
got_df <-map_df(got_chars, `[`, 
                c("name", "culture", "gender", "id", "alive"))
head(got_df)
```
Note: Keep an eye on your data types. When in doubt, use simpler `map` functions to build your final dataframe.

##Safely
Troubleshooting vectorized functions can be a pain.
```{r safely}
numbers <- c(1:10)
map_dbl(numbers, sqrt)
```  
Pretend this is actually 1.5 million data points: incorrect values aren't going to be something you can eyeball.  
```{r safely2, error = TRUE}
numbers_bad <- c(1, 2, 3, 4, 5, "Tyrion Lannister",7, 8, 9)
map_dbl(numbers_bad, sqrt)

```
##Safely
Purrr adds a family of functions under `safely()` that make troubleshooting these a breeze. 

1.`safely` returns a list with `result` and `error`  
2.`quietly` returns a list with `result`,`output`, `messages`, and `warnings`    
3.`possibly` will fill any errors with a specified value.

##Safely
```{r safely4}
numbers_bad <- list(1,"Tyrion Lannister",3)
safe_sqrt <- safely(sqrt)

map(numbers_bad, safe_sqrt)
```
##Possibly
```{r possibly}
numbers_bad <- list(1,"Tyrion Lannister",3)
possible_sqrt <- possibly(sqrt, otherwise = NA)

map_dbl(numbers_bad, possible_sqrt)
```

##loading data
```{r data}
library(readr)
#make a list of all files in a directory
files <- list.files("./data/mtls/")
#add file path info
files <- paste0("./data/mtls/", files)
#map it!
data <- map(files, read_table )
data[[1]]
```
##Multiple Models
lets take the gapminder dataset. First we have to structure the data. 
```{r gapminder}
library(gapminder)
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()
#sanity check it's still all there
head(by_country$data[[1]])
```

##Multiple Models
Our model. Note that we're wrapping this inside a function so that it plays nicely with `map`
```{r model}
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}
```
```{r fit}
by_country <- by_country %>% 
  #now add a column that contains the fit models
  mutate(model = map(data, country_model))
head(by_country, n = 3)
```  
This is awesome! It's still a dataframe, so all of our `dplyr`, `tidyr`, and `purrr` magic works just fine!  

##Multiple Models
Challenge: Add a column to this dataframe that contains the residuals.

##Multiple Models
Challenge: Add a column to this dataframe that contains the residuals.
```{r challenge answer}
by_country <- by_country %>% 
  mutate(
    resids = map2(data, model, add_residuals)
  )
```

##Multiple Models
Now we just have to unnest, and you're off to the races!
Note: the nested dataframe is the most useful base to work from. create task oriented dataframes for things like plotting residuals.
```{r unnest}
#in this case, we're dropping our models
resids <- unnest(by_country, resids)
```
##Multiple Models
Lets take a look at our residuals.
```{r plot residuals}
resids %>% 
  ggplot(aes(year, resid, group = country)) +
    geom_line(alpha = 1 / 3) + 
    facet_wrap(~continent)
```
