---
title: "Mapmaker, Mapmaker, Make me a Map!"
author: "Intro to Data Science for Biology"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Intro 
For this week's homework, let's work on mapping the covid-19 data. You have two choices of data source. The first is the coronavirus data we have already loaded.

```{r corona}
library(coronavirus)
head(coronavirus)
```

The second is a newer dataset. It harvests data that is from the New York Times. It is focused solely on the US. To install it, you'll need to do the following

```{r, eval = FALSE}
#if you don't have it already
install.packages("devtools")

#install the library from github
devtools::install_github("covid19R/covid19nytimes")
```

```{r nytimes}
library(covid19nytimes)
covid_states <- refresh_covid19nytimes_states()

head(covid_states)

covid_counties <- refresh_covid19nytimes_counties()

head(covid_counties)

```


Now, you have three data sets to choose from! Countries, states, or counties. Remember, with the coronavirus data, you have to do some dplyr::summarizing to get it down to countries, though!

## Maps to use for this assignment

OK, so, we need world, US state, and US county maps - depending on which of the three datasets you chose
```{r}
library(sf)

#The world
library(rnaturalearth)
world_map <- ne_countries()

#US States
library(USAboundaries)
us_states <- us_states()

#US Counties
us_counties <- us_counties()
```

Armed with this, let's make some maps!

### Questions

1. Which data set - or aspect of a single data set, are you most interested in? Sort through the datasets. What is there? Is it the world? A single country? Multiple contries? All states? Counties in one state? 

Filter or summarize your data to just what you are interested in, in terms of space.

For example
```{r}
library(dplyr)
florida_covid <- covid_counties %>%
  filter(stringr::str_detect(location, "[Ff]lorida"))

florida_map <- us_counties %>%
  filter(state_name == "Florida")
```

2. What type or types of data from that dataset are you interested in? Why? Filter the dataset to that data type only.

3. What do you want to learn from this slice of the data? Formulate a question and write it out here.

4. Filter and manipulate the data so that it is in a format to be used to answer the question.

5. Join the covid data with spatial data to build a map.

6. Create a map from this data! Make it awesome!

7. What do you learn from the map you made?

8. This static map is, I'm sure, great. Load up `tmap` and make it dynamic! Is there anything different you can learn from this form of visualization?

9. Last... it's time to start thinking about your final project. Name the one or more data sources you are interested in exploring. Tell us a bit about what is in them.
