---
title: "Mapmaker, Mapmaker, Make me a Map!"
author: "Intro to Data Science for Biology"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Intro 
For this week's homework, we're going to focus on making maps with the TB incidence data from the World Health Organization and a little bit of Sea Surface Temperature.

First, the TB data, which is at the country level. Let's take a look

```{r who_load, message=FALSE}
library(readr)

tb_world <- read_csv("../Labs/data/who_tb_data.csv")

tb_world

```

We can then use this worh the world map from ggplot2
```{r world, message=FALSE}
library(tidyverse)
library(ggplot2)

#get a map
worldmap <- map_data("world")
```

### Joins and Mapping

1. Explore joining these two data sets. What does not join well? Can you fix any of them via changes in the tb or worldmap dataset? Make whatever fixes you can, and then join them with tools from `dplyr` for....  

2. Using ggplot2, make a map showing change in mortality over four years (filter to four years, and then use `facet_wrap`).


### Shapefiles

Let's work with a shapefile of international borders now. Note the libraries I'm loading - you should in your homework, too!

```{r borders, message=FALSE}
library(rgdal)
library(sp)
library(broom)

world_shapefile <- readOGR("../Labs/data/TM_WORLD_BORDERS_SIMPL-0.3")
```

What's inside?
```{r show_borders}
plot(world_shapefile)

head(world_shapefile@data)
```

Oh neat! Notice the ISO2 and ISO3 columns? Those are designed to match the standardized columns from the `tb_world` data! Once we turn this into a data frame, should be a snap to merge! Let's make the same plot as we did previously, but now with shapefile.


So, TB in 2015!

```{r first_fade_shapefile}
#get a map
worldmap_shp <- tidy(world_shapefile, region="NAME")

worldmap_shp_df <- left_join(worldmap_shp, world_shapefile@data,
                          by = c("id" = "NAME")) %>%
  rename(iso2 = ISO2, iso3 = ISO3)

#filter to 2015
tb_2015 <- tb_world %>% filter(year == 2015)

#join
incidence_df <- left_join(worldmap_shp_df, tb_2015)

#plot
ggplot(incidence_df, aes(x = long, y = lat, 
                         group = group, fill=est_incidences)) +
  geom_polygon() +
  scale_fill_gradient(low = "blue", high = "red")


```


OK, now, some faded examples using the above as a starting point

3. Mortality in 2000

```{r eval=FALSE}
#get a map
worldmap_shp <- tidy(world_shapefile, region="NAME")

worldmap_shp_df <- ____(worldmap_shp, world_shapefile@data,
                          by = c("id" = "NAME")) %>%
  rename(iso2 = ISO2, iso3 = ISO3)

#filter to 2000
tb_2000 <- tb_world %>% filter(year == ____)

#join
incidence_df_2000 <- left_join(____, ____)

#plot
ggplot(incidence_df_2000, aes(x = ____, y = ____, 
                         group = group, fill=est_mortality)) +
  geom_polygon() +
  scale_fill_gradient(____)
```

4. Incidence per 100K, but, only by interval, in 2016 - feel free to do something difference with the fill
```{r, eval=FALSE}
#get a map
worldmap_shp <- ____(____, ____="NAME")

worldmap_shp_df <- ____(____, world_shapefile@____,
                          by = c("id" = "NAME")) %>%
  rename(iso2 = ISO2, iso3 = ISO3)

#filter to 2016
tb_2016 <- tb_world %>% ____(year == ____)

#join
incidence_df_2016 <- left_join(____, ____, by = c("region" = "country"))

#plot
ggplot(incidence_df_2016, aes(x = ____, y = ____, 
                         group = ____, 
                         fill=____(est_incidences_per_100K_people,5))) +
  ____() +
  scale_fill_brewer(____ = "YlOrBr",
                    ____ = guide_legend("Incidences per 100K"))
```


5. Make *two* maps that compare total incidence to incidence per 100K averaged across all years. So, there are three steps - first, the averaging. Then, the joining. Then, the plotting.  


### Leaflet, Rasters, Shapefiles, and Putting it All Together

6. Let's take a look at the distribution of global SST anomalies. I've grabbed last week's data from http://www.cpc.ncep.noaa.gov/products/GIS/GIS_DATA/sst_oiv2/sst-anom.php (download it from this site [here](http://biol355.githib.io/Lectures/data/sst_anom.bin.20180318.tif) and I'd like you to load it, and the [marine ecoregions of the world](http://biol355.githib.io/Lectures/data//data/MEOW-TNC.zip) in order to make a map where we can explore how that temperature anomoly varies by ecoregion. This means you will be layering polygons from a shapefile on top of a raster using leaflet. Feel free to play with color palettes and any other property you'd care to examine!

Note, you'll need to supply the polygons name as the "data" to `leaflet` and supply the raster separately. `fillOpacity = 0` will also make them transparent

Also, include the following line to take out land, etc.
```{r, eval=FALSE}
values(sst)[values(sst)>100] <- NA
```

This sets all of those big values to NA

It should look something like below - but use whatever color palette or strategy that you think will make things pop the most! Feel free to also learn more about using leaflet at https://rstudio.github.io/leaflet/

```{r test, echo=FALSE, message=FALSE, cache=TRUE}
library(raster)
library(leaflet)

sst <- raster("../Lectures/data/sst_anom.bin.20180318.tif")
values(sst)[values(sst)>100] <- NA

meow <- readOGR("../Labs/data/MEOW-TNC")

sstpal <- colorNumeric("RdBu", 
                    values(sst),
                    na.color = "transparent")

leaflet(meow) %>%
  addRasterImage(sst, colors = sstpal) %>%
  addPolygons(weight = 1, color="black", fillOpacity = 0)%>%
   addLegend(pal = sstpal, values = values(sst))

```


Based on the map **YOU** make, what do you see that is of interest?