#Data cleanup with Tidyr
*Due March 22 2018*

###Warmup with some Faded Examples!  
Grab `Sale_Prices_City.csv` and bring it in to R, [link to data here](../Data/Sale_Prices_City.csv).

First, convert it from long to wide, with a column for each year/month
```{r gather, error = TRUE, eval=FALSE}
sales_long <- sales %>%
    ______(___ = time_point, 
           _____ = sale_price, 
           -_________, -_________, -________, -________)
```


Drop the NAs
```{r drop NA, error = TRUE, eval=FALSE}
sales_long <- sales_long %>%
    ______(______(sale_price))
```

Split up year and month into two columns
```{r split date, error = TRUE, eval=FALSE}
sales_long <- sales_long %>%
    ________(time_point, c("year", ______), ___ = "-")
```

Make the following string:
```{r make string}
my_string <- "Some people, when confronted with a problem, think 'I know, I'll use regular expressions.' Now they have two problems."
```
Make it all uppercase
```{r uppercase, error = TRUE, eval=FALSE}
__________(my_string)
```
Remove all instances of the letter e
```{r goodbyeee, error = TRUE, eval=FALSE}
______________(my_string, ______ = "e", replacement = ___)
```
Split this string into a vector of individual words
```{r split TRUE, eval=FALSE}
split_string <- __________(my_string, pattern = ___, simplify = ______)
```


Find the words that start with consonants.
```{r TRUE, eval=FALSE}
str_view(split_string, "___________")

```

###Exercise 1
Load up the raw hadcrut data, [link here](../Data/hadcrut_temp_anomoly_wide_1850_2015.csv). We've been using this in a long format, but it actually is supplied as wide data. Use your skills with Tidyr to make it look like the long data we've been using in class!

##Please submit the following as an RMD as well as the knitted HTML:

Faded Examples: One example per chunk.

Excercise 1: Code required for loading data, converting to long format, and then use `head()` to display the first five lines.Be sure to include steps where you check your work with `str` and and explain the relavent parts of what you see! Feel free to do this in comments. an example of this could be:

```{r}
#First I will create a vector of 100 random numbers between 0 and 1 using runif
x <- runif(100)
#I will then use mean to find the average of this vector
mean(x)
```

I want to know that you know what your code is doing!
