#Data cleanup with Tidyr
*Due March 22 2018*

###Warmup with some Faded Examples!  
Grab `Sale_Prices_City.csv` and bring it in to R, [link to data here](../Data/Sale_Prices_City.csv).

First, convert it from long to wide, with a column for each year/month
```{r gather, error = TRUE, eval=FALSE}
sales_long <- sales %>%
    ______(___ = time_point, 
           _____ = sale_price, 
           -_________, -_________, -________, -________)
```


Drop the NAs
```{r drop NA, error = TRUE, eval=FALSE}
sales_long <- sales_long %>%
    ______(______(sale_price))
```

Split up year and month into two columns
```{r split date, error = TRUE, eval=FALSE}
sales_long <- sales_long %>%
    ________(time_point, c("year", ______), ___ = "-")
```

Make the following string:
```{r make string}
my_string <- "Some people, when confronted with a problem, think 'I know, I'll use regular expressions.' Now they have two problems."
```
Make it all uppercase
```{r uppercase, error = TRUE, eval=FALSE}
__________(my_string)
```
Remove all instances of the letter e
```{r goodbyeee, error = TRUE, eval=FALSE}
______________(my_string, ______ = "e", replacement = ___)
```
Split this string into a vector of individual words
```{r split TRUE, eval=FALSE}
split_string <- __________(my_string, pattern = ___, simplify = ______)
```


Find the words that start with consonants.
```{r TRUE, eval=FALSE}
str_view(split_string, "___________")

```

###Exercise 1
Load up the raw hadcrut data, [link here](../Data/hadcrut_temp_anomoly_wide_1850_2015.csv). We've been using this in a long format, but it actually is supplied as wide data. Use your skills with Tidyr to make it look like the long data we've been using in class!

###Exercise 2
Download the messy weather dataset from class today (3/7/2018) [link here](../Data/weather_messy.csv). This is a dataset that contains some weather measurements but structured in an...interesting way. Go through the cleanup steps required to go from this raw data to a cleaned up dataframe in the long format. This dataframe should have separate columns for year, month, day, and each measurement type. 

**Suggested workflow:**  
1. Read in data.   
2. Deal with any extraneous blank spaces and redundant columns.  
3. Fix column names (Hint:beyond the first three columns, each column represents a day of the month. )  
4. Fill in missing year/months.  
5. OPTIONAL: Combine date info into one column (you don't need to do this, but it might help you reduce the load on your brain and better grock what's happening).  
6. Start playing with wide vs long. HINT: This dataset is one of those that we mentioned which has characteristics of both long and wide data: both `gather` and `spread` might be needed to get it to its final form.  
7. Deal with NAs and other data-entry issues. HINT: to avoid deleting most of the datset, let NA values remain in the "events" column.  
8. Final tweaks: make sure numbers are numeric and characters are characters.  
9. If you combined date columns before, separate them back out again so you have one for year, one for month, and one for day. 
10. Give me a plot, any plot! Be sure to make it look nice!


**Function Toolbox**
If you get stuck, consider using the following functions:
`gather`, `spread`, `unite`, `separate`, `vis_dat`, `str`, `filter`, `select`, `names`, and `fill`. Remember, if you want to know more about a function just type it into the console with a ? in front.  

##Please submit the following as an RMD as well as the knitted HTML:

Faded Examples: One example per chunk.

Excercise 1: Code required for loading data, converting to long format, and then use `head()` to display the first five lines

Excercise 2: Because we are going over the bulk of this workflow in class, please submit a well-formatted write up in which each step is explained in english. Be sure to include steps where you check your work with `str` and `vis_dat` and explain the relavent parts of what you see! Feel free to do this in comments. an example of this could be:

```{r}
#First I will create a vector of 100 random numbers between 0 and 1 using runif
x <- runif(100)
#I will then use mean to find the average of this vector
mean(x)
```

I want to know that you know what your code is doing!
