---
title: "Spatial Coronavirus"
author: "Biol355"
date: "3/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The world!

For today, I've given you a number of shapefiles associated with worldmaps, state maps of the US, and county maps of Massachusetts and the coronavirus. You can download the data [here](./Data/coronavirus_shapefiles.zip). I'd like you to take a gander and think about what you can do with them.

To start with, let's take a look at the world.

```{r world, warning=FALSE, message=FALSE}
library(sf)
library(ggplot2)

worldmap <- st_read("./data/coronavirus_shapefiles/worldmap.shp")

ggplot(worldmap) +
  geom_sf() +
  coord_sf()
```

That's a nice example workflow for loading and plotting one single shapefile. The shapefiles I have given you have multiple rows for dates, geospatial areas, and different case types (confirmed, death, recovered). You should be somewhat familiar with this already.

## Lab Exercises

With that in mind, I'd like you to - 

1. Plot the spatial distribution of coronavirus cases.  Color by date.  
\
2. Show the distribution of mortality across the world by country using polygons.  
\
3. Within the US, show the total confirmed cases of coronavirus across the country by state. Feel free to use points or polygons here.   
\
4. Show the progression of conformed coronavirus in Massachusetts spatially over time using facets and dates.  
\
5. Take the map above you liked the most and reproject it! Do something cool and funky looking!
