---
title: "Joins and Geospatial Data"
author: "Biol 355: Intro to Data Science for Biology"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

### Map maker, map maper, make me a map!

Today we're going to look at how we can use joins and geospatial data more explicitly to make maps. Maps are among the first data visualizations that ever occured. And some of the most powerful. They're also one of the places where joins become incredibly important, as to put data on a map we have to join our data with a geospatial description of the map we want.

### Death from Heart Disease

Today's data set that we'll be using is a data set of heart disease mortality from the CDC

```{r}
#load the data and prep
#for some data manipulation
library(readxl)
library(dplyr)

heart_disease <- read_excel("./data/hd_all.xlsx", 
                            na="Insufficient Data")

head(heart_disease)
```

OK, we see that we have state, county, and information on death. FIPS codes, FYI, are standardized county codes. They'll become important in a bit.

### Introducing: Maps

There are a LOT of ways to get map data into R. We're going to begin with the simplest - grabbing it from an R package. The `USAbouncaries` package has boundaries of states and counties in the US through time. To learn more, check out https://lincolnmullen.com/software/usaboundaries/. It has a wonderful set of county maps that are easy to use with the `us_counties()` function! 

```{r}
#install the USAboundaries library if you
#don't have it
#devtools::install_github("ropensci/USAboundaries")
#devtools::install_github("ropensci/USAboundariesData")
library(USAboundaries) 
library(sf)

map_sf <- us_counties()

```

Nice! So, what is in here?

```{r}
class(map_sf)

st_crs(map_sf)
```

OK, it is indeed an `sf` object.

```{r}
map_sf
```

OK! there is a kit tgere! Most notably, I want you to see that while we have county names in `name` and state names in `state_name` we also have these fp codes - known as FIPS codes. There are Federal Information Processing Standards codes used by the government, as names are... flexible. For more, see https://en.wikipedia.org/wiki/FIPS_county_code - we'll talk about them in a bit.

To make sure all is well, let's take a gander!

```{r}
library(ggplot2)
ggplot(data=map_sf) +
  geom_sf(fill = "grey") +
  coord_sf() +
  theme_minimal()
```

Great!

### Getting ready to join maps and data

OK, so, we want to see how death from heart disease varies by county! To do this, we'll need to join the data frame and sf object. 

First, we need to fix up the columns in our `map_sf` data set to state and county names.

```{r}
library(tidyr)
library(stringr)

map_sf <- map_sf %>%
  rename(County = name, 
         State = state_name)
```

Now let's test out the join! Before we do the big join, let's look at mismatch. This is the perfect job for `anti_join`

```{r}
bad_match <- anti_join(heart_disease, map_sf)

nrow(bad_match)
```

Hey, not bad! 18! Let's see what the problems are

```{r}
bad_match[,1:2]
```

So - some of it is cities rather than counties. Understandable. Someo f it is the USVIs. And some of it might be simply due to funny names. Let's look at one - that Prince of Wales county in Alaska as a case study.

```{r}
bad_match[1,]

map_sf %>%
  filter(State=="Alaska") %>%
  filter(stringr::str_detect(County, "Prince")) %>%
  as_tibble()  %>%
  select(State, County) 
```

Oh! So, names are off. Huh. What if instead of joining by state and county we used something else - like a standard FIPS code! We have a small problem - the `map_sf` object needs a full state-county FIPS code. And then we can make the join ignore state and county using the `by` argument to any join.

```{r}
map_sf <- map_sf %>%
  mutate(FIPS_Code = stringr::str_c(statefp, countyfp) %>% as.numeric)


anti_join(heart_disease, map_sf,
          by = "FIPS_Code")
```


Only 9. Were we trying to make this perfect, we'd try and figure out why, but, for now, let's soldier on. 9 is acceptable.

What if we go the other way?

```{r}

anti_join(map_sf, heart_disease,
          by = "FIPS_Code")
```

OH! Territories. There's no data for them. That's OK for now.

### Joining maps and data

So we know that there are some territories and cities left in the heart disease data, and we don't want to worry about them for the moment. Given that we've got some mismatch on both sides, we want to use an `left_join` - although we could also `full_join` to eliminate some information. Try it both ways.

Let's also get rid of Hawaii and Alaska for the moment, as well as anything that is a territory. So, just the lower 48!

We will also need to make sure that this is indeed an sf object. Joining spatial and non-spatial objects leads to non-spatial objects!

```{r}
heart_disease_map_data <- left_join(heart_disease, map_sf) %>%
  filter(jurisdiction_type != "territory") %>%
  filter(!(State %in% c("Hawaii", "Alaska"))) %>%
  st_as_sf(crs = 4326)
```

And now - let's plot it!

```{r}
heart_map <- ggplot(data=heart_disease_map_data, 
       mapping = aes(fill = Death_Rate)) +
  geom_sf(color = NA) +
  theme_minimal()

heart_map
```

### Making pretty choropleths

So, this is a choropleth map. There are some issues - the scale is hard to resolve, and everything is hard to see. There are a lot of ways we could rectify it. Here are a few tricks.

First, a better scale. Maybe a gradient?

```{r}
heart_map +
  scale_fill_gradient(low = "white", high = "red")
```

OK, not bad! Not great, but much better. Are there gradients you'd prefer?

However, the bigger problem is that we have a huge range to cover. One common way to make choropleths is to bin data into categories and go from there. Binning in combination with a nice color scale - say via Color Brewer  - cab be great. Remember Color Brewer? http://colorbrewer2.org/ actually uses maps as it's demo! And there's a `scale_fill_brewer` function in ggplot2.

So let's make bins and plot that instead!

```{r}
heart_map_binned <- ggplot(data=heart_disease_map_data, 
       mapping = aes(fill = cut_interval(Death_Rate, 5))) +
  geom_sf(color = NA) +
  theme_minimal()

heart_map_binned +
    scale_fill_brewer(palette="Reds")
```

MUCH nicer. Now we can really start to see some hot spots.

### Faded Examples
For our faded examples, today we're going to look at data on TB incidence from the World Health organization. This look at TB for the entire planet at the country level. Let's take a look

```{r who_load, message=FALSE}
library(readr)

tb_world <- read_csv("./data/who_tb_data.csv")

tb_world

```

There's a lot going on here - in particular let's focus on estimated insidences, incidences per 100K people, and mortality. `upr` and `lwr` denote upper and lower bounds to estimates.

So, let's make a simple map of 2015.

Before we do that, I want to introduce you to `rnaturalearth`. It's a package that I use almost daily for really nice maps. While it doesn't have detailed county level data, it has a ton of data at the world and state/province level coupled with iso3  standardized international codes. Let's see it in action for just ONE type of map.

```{r rnaturalearth}
library(rnaturalearth)

worldmap <- ne_countries(returnclass = "sf")

names(worldmap)

```

WHOAH! That is a ton of information. However, I'd like you to notice the iso_a3 code, analagous to the iso3 code in the WHO data. We can use it to join on.

So, 2015 - note, I'm going to left join on the worldmap in order to ensure that we retain all of the countries of the world

```{r first_fade}
#get a map
worldmap <- ne_countries(returnclass = "sf")

#filter to 2015
tb_2015 <- tb_world %>% filter(year == 2015)

#join
incidence_df <- left_join(worldmap, tb_2015, by = c("iso_a3" = "iso3"))

#plot
ggplot(incidence_df, aes(fill=est_incidences)) +
  geom_sf() +
  scale_fill_gradient(low = "blue", high = "red")


```

Nice!

Let's look at mortality in 2000

```{r eval=FALSE}
#get a map
worldmap <- ne_countries(returnclass = "____")


#filter to 2015
tb_2000 <- tb_world %>% filter(year == ____)

#join
incidence_df_2000 <- left_join(worldmap, ____, by = c("iso_a3" = "iso3"))

#plot
ggplot(incidence_df_2000, aes(fill=est_mortality)) +
  geom_sf() +
  scale_fill_gradient(____)
```

```{r second_fade, eval=FALSE, echo=FALSE}
#get a map
worldmap <- ne_countries(returnclass = "sf")


#filter to 2000
tb_2000 <- tb_world %>% filter(year == 2000)

#join
incidence_df_2000 <- left_join(worldmap, tb_2000, by = c("iso_a3" = "iso3"))

#plot
ggplot(incidence_df_2000, aes(fill=est_mortality)) +
  geom_sf() +
  scale_fill_gradient(low = "blue", high = "red")
```

Now let's look at incidence per 100K, but, only by interval, in 2016 - feel free to do something difference with the fill

```{r, eval=FALSE}
#get a map
worldmap <- ____(returnclass = "____")


#filter to 2016
tb_2016 <- tb_world %>% ____(year == ____)

#join
incidence_df_2016 <- left_join(worldmap, ____, by = c("iso_a3" = "iso3"))

#plot
ggplot(incidence_df_2016, aes(fill=____(est_incidences_per_100K_people,5))) +
  ____() +
  scale_fill_brewer(palette = "YlOrBr",
                    guide = guide_legend("Incidences per 100K"))
```


```{r third_fade, include = FALSE, echo=FALSE}
#get a map
worldmap <- ne_countries(returnclass = "sf")


#filter to 2016
tb_2016 <- tb_world %>% filter(year == 2016)

#join
incidence_df_2016 <- left_join(worldmap, tb_2016, by = c("iso_a3" = "iso3"))

#plot
ggplot(incidence_df_2016, aes(fill=cut_interval(est_incidences_per_100K_people,5))) +
  geom_sf() +
  scale_fill_brewer(palette = "YlOrBr",
                    guide = guide_legend("Incidences per 100K"))
```

### Pair Programming Exercises

1. What does not join well? Can you fix any of them via changes in the tb or worldmap dataset? Or should you not bother?   

2. Can you make a map showing change in mortality over four years (filter to four years, and then use `facet_wrap`)?


### Wibbly Wobbly Visualizations

This is all well and good, but what if you want to be google maps? There are a few packages out there. [leaflet](https://rstudio.github.io/leaflet/) and [tmap](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html) are the dominant. I <3 leaflet, but I find tmap is easier to get used to, as it functions similarly to ggplot. There's also a nice primer in [Geocomputation in R](https://geocompr.robinlovelace.net/adv-map.html#interactive-maps). Both are great, but, for the moment, let's try `tmap`!

Let's start with this heart diesease data to see how it works in general. The basic way that tmap works is that it uses `tmap_shape()` to reference a dataset, and then a series of functions after to add features from that dataset - e.g., `tm_polygons()`. It doesn't use NSE, so we put quotes around column names. 

```{r tmap}
library(tmap)

first_tmap <- tm_shape(shp = heart_disease_map_data) +
  tm_polygons(col = "Death_Rate")
```

Hey, wait, I said interactive! To make it so, we call `tmap_mode("view")` before we run things. Right now we're in the `plot` view.

```{r}
tmap_mode("view")

first_tmap
```

OH! Cool! Play around a bit here!

You can still do a lot with `tmap` and the [vignette](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html) shows off a lot like facets. Also, see how we add palettes. I'm using a list because we have two elements. And see `tmaptools::palette_explorer()` for more on colors.

```{r}
tm_shape(incidence_df_2016) +
  tm_polygons(c("est_incidences", "est_mortality"),
              palette = list("magma", "viridis")) +
  tm_facets(sync = TRUE, ncol = 2)
```

We can also use facets in a non-interactive way that is quite useful..

```{r}
tmap_mode("plot")
tm_shape(incidence_df_2016) +
  tm_polygons("est_incidences") +
  tm_facets(by = "region_un")
```

### Exercise

Turn the following two examples:

```{r, eval = FALSE}
ggplot(incidence_df, aes(fill=est_incidences)) +
  geom_sf() +
  scale_fill_gradient(low = "blue", high = "red")

ggplot(incidence_df_2016, aes(fill=cut_interval(est_incidences_per_100K_people,5))) +
  geom_sf() +
  scale_fill_brewer(palette = "YlOrBr",
                    guide = guide_legend("Incidences per 100K"))
```

into tmap syntax using `tm_shape()`, `tm_polygons()`, and the `palette` argument. Explore palettes if need be. To change the legend name, see `?tm_polygons` for the proper argument. Make them interactive!

### Final Lab Exercise!

OK, this is all well and good. Working alone or in groups, I'd like you to load up the `coronavirus` package. Remembering that
a. Each row represents one report, 
b. Each report has a type (confirmed, death, or recovered), 
c. Each report comes with  other geospatial information - country, state, etc.

I'd like you to visualize some aspect of the spatial distribution of coronavirus. The maximum date in the dataset isn't current, so, just remember that. If you want to look at a more current dataset, check https://data.humdata.org/dataset/novel-coronavirus-2019-ncov-cases - but you will have to look hard at the data to understand what is there. (I'm working on a method to get something more current within R along with Rami Krispin and others.)

```{r corona, echo = FALSE}
library(coronavirus)

cv_max <- coronavirus  %>%
    mutate(Country.Region = gsub("United States of America", "United States", Country.Region)) %>%
  group_by(Country.Region, type) %>%
  summarize(cases = sum(cases))


cv_max_map <- left_join(cv_max, worldmap, by = c("Country.Region" = "name")) %>%
  tidyr::pivot_wider(names_from = type.x, values_from = cases) %>%
  st_as_sf(crs=4326) %>%
  mutate(`log10 Cases` = log10(confirmed),
         `Percent Mortality` = death/confirmed*100,
         `Percent Recovered` = recovered/confirmed*100)

tmap_mode("view")
tm_shape(cv_max_map) +
  tm_polygons(c("log10 Cases", "Percent Mortality", "Percent Recovered"),
              palette = list("viridis", "RdYlBu", "YlGn"),
              free.scales = TRUE) +
  tm_facets(ncol = 3, sync = TRUE)

```